---
title: "BACT: Bayesian Adaptive Clinical Trial"
author: "Thevaa Chandereng, Donald Musgrove, Tarek Haddad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bayesian trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


# Introduction


Clinical Trial is randomized intervention comparing a new treatment/drug to a current treatment/drug (control). 
Most of the analysis developed for clinical trials have been highly focused on frequentist approach. 
The Bayesian adaptive trial has not gained much popularity until recent years. 

Bayesian adaptive trials provide much more flexibility over the conventional frequentist approach in terms of the design and analysis of a clinical trial.
Bayesian adaptive trials can incorporate historical data which helps reduce the sample size required in a clinical trial. 
Bayesian adaptive trials are also useful in an interim analysis for early stopping for success or futility. 
Bayesian adaptive trials have also been used in multiple treatments vs control which allows dropping a treatment which is not successful during a trial. 
Statistician has altered randomization ratio to promote efficiency in a trial or even to increase recruitment of patients into a trial. 

However, simulation of a Bayesian adaptive trial is much more complex compared to the general frequentist approach. 
This package should aid with the simulation of Bayesian adaptive trials. 


If you use BACT in published research, please cite:



If after reading through this vignette you have questions or problems
using LPWC, please post them to https://github.com/thevaachandereng/BACT/issues.
This will notify the package maintainers and benefit other users.


You should **not** email your question to the package authors, as we will
just reply that the question should be posted to the
**Github support site**.



# Running BACT

Prior to analyzing your data, the R package needs to be installed.

The easiest way to install BACT is through CRAN:
```{r, eval = FALSE}
install.packages("BACT")
```


There are other additional ways to download BACT.
The first option is most useful if want to download a specific version of BACT
(which can be found at https://github.com/thevaachandereng/BACT/releases).
```{r, eval = FALSE}
devtools::install_github("thevaachandereng/BACT@vx.xx.x")
#OR 
devtools::install_version("BACT", version = "x.x.x", repos = "http://cran.us.r-project.org")
```


The second option is to download through GitHub. 

```{r, results='asis', eval=FALSE}
devtools::install_github("thevaachandereng/BACT")
```


After successful installation, the package must be loaded into the working space:

```{r lib, results="asis", eval=TRUE}
library(BACT)
```



# Recruitment

The recruitment of patients are assumed to follow a poisson process. 


