---
title: "BACT: Bayesian Adaptive Clinical Trial"
author: "Thevaa Chandereng, Donald Musgrove, Tarek Haddad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bayesian trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


# Introduction

Clinical Trial are randomized intervention comparing a new treatment/drug to a current treatment/drug (control). 
Most analysis developed for clinical trials have been highly focused on frequentist approach. 
Bayesian adaptive trial have not gain much popularity until recent years. 

Bayesian adaptive trials provides much more flexibility over the conventional frequestist approach in terms of the design and analysis of a clinical trial.
Bayesian adaptive trials 



If you use BACT in published research, please cite our preprint:


If after reading through this vignette you have questions or problems
using LPWC, please post them to https://github.com/thevaachandereng/BACT/issues.
This will notify the package maintainers and benefit other users.


You should **not** email your question to the package authors, as we will
just reply that the question should be posted to the
**Github support site**.



# Running BACT

Prior to analyzing your data, the R package needs to be installed.

The easiest way to install BACT is through CRAN:
```{r, eval = FALSE}
install.packages("BACT")
```


There are other additional ways to download BACT.
The first option is most useful if want to download a specific version of BACT
(which can be found at https://github.com/thevaachandereng/BACT/releases).
```{r, eval = FALSE}
devtools::install_github("thevaachandereng/BACT@vx.xx.x")
#OR 
devtools::install_version("BACT", version = "x.x.x", repos = "http://cran.us.r-project.org")
```


The second option is to download through GitHub. 

```{r, results='asis', eval=FALSE}
devtools::install_github("thevaachandereng/BACT")
```


After successful installation, the package must be loaded into the working space:

```{r lib, results="asis", eval=TRUE}
library(BACT)
```


