---
title: "bayesCT: An R package for Simulation in Adaptive Bayesian Clinical Trials"
author: "Thevaa Chandereng, Graeme Hickey, Donald Musgrove, Tarek Haddad, Timothy Hanson, Theodore Lystig"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{bayesian trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


# Introduction

Randomized controlled trials (RCT) are the gold standard of clinical studies of medical and pharmaceutical devices. In an RCT, subjects are randomized to an intervention arm, with the standard framework being the comparison of a new treatment/drug to a current treatment/drug (control). The majority of research in the design of RCTs and their application has been based on the frequentist paradigm. In recent years, the adaptive Bayesian trial design approach has gained attention.

Adaptive Bayesian trials provide much more flexibility over conventional frequentist approaches in terms of the design and analysis of a clinical trial. Notably, adaptive Bayesian trials:

- can incorporate historical data, which helps reduce the sample size required;
- can be exploited for interim analyses of early stopping for success or futility;
- can be used in trials with multiple treatment arms (versus a single control), allowing for treatments not demonstrating success to be dropped; and
- can be used to alter the randomization ratio to improve trial efficiency or even to increase trial recruitment.

A fundamental requirement of an adaptive trial is the *a priori* and interim evaluation of the operating characteristics. In a adaptive trials, these are generally not of closed-form; hence, simulation approaches are required. Simulation of adaptive Bayesian trial designs have historically been substantially more complex compared to frequentist approaches. This package removes this obstacle and allows for simulation of adaptive Bayesian trials under a range of designs and outcome types.

If you use `bayesCT` in published research, please cite: **TBA**.

If after reading through this vignette you have questions or problems using `bayesCT`, please post them to https://github.com/thevaachandereng/bayesCT/issues. This will notify the package maintainers and benefit other users. Please note that comments and questions should **not** be emailed directly to the package authors.

# Running bayesCT

Prior to analyzing your data, the R package needs to be installed.

The easiest way to install `bayesCT` is through CRAN:

```{r, eval = FALSE}
install.packages("bayesCT")
```

There are additional ways to download `bayesCT`. The first option is most useful if want to download a specific version of `bayesCT` (which can be found at https://github.com/thevaachandereng/bayesCT/releases):

```{r, eval = FALSE}
devtools::install_github("thevaachandereng/bayesCT@vx.xx.x")

# or 

devtools::install_version("bayesCT", version = "x.x.x", repos = "http://cran.us.r-project.org")
```

The second option is to download the most recent stable version through GitHub:

```{r, message = FALSE, prompt=FALSE}
devtools::install_github("thevaachandereng/bayesCT")
```

After successful installation, the package must be loaded into the working space:

```{r lib, results="asis", eval=TRUE}
library(bayesCT)
```


# Recruitment

The recruitment of subjects is assumed to follow a Poisson process. Since trial recruitment is independent, the ["memoryless property"](https://en.wikipedia.org/wiki/Memorylessness) modeling of subject recruitment. However, we cannot assume a homogeneous Poisson process since subject recruitment varies over time. Thus, the `enrollment()` function assumes a piecewise stationary Poisson process with different $\lambda$ parameters over different time intervals. The function also omits an enrollment date considering time zero as study initiation. 

For instance, if we have the following piecewise function of $\lambda$, with $t$ representing the time interval,  

\[  \lambda = \left\{
\begin{array}{ll}
      0.3 & t \in [0, 5) \\
      0.7 & t \in [5, 10) \\
      0.9 & t \in [10, 15) \\
      1.2 & t \in [15, \infty) \\
\end{array} 
\right. \]

then to simulate individual patient enrollment dates with a sample size (`N_total`) of 50, one would use

```{r}
enrollment(param = c(0.3, 0.7, 0.9, 1.2), 50, time = c(5, 10, 15))
```


# Randomization scheme

The randomization allocation is important component in the design of an RCT. Complete randomization is not always ideal due to the chance of drawing a large block of a single treatment arm, which can impact the time to enrollment completion. Therefore, a block randomization allocation is generally preferable. Block randomization allocation allows for different randomization ratios but they must be in integer form. In addition, the block size should also be an integer that is divisible by the sum of randomization allocation. 

For instance,

```{r, eval = FALSE}
randomization(N_total = 140, block = 7, allocation = c(2, 1))
```

will not work because the sum of the allocation is not a multiple of the block. 

```{r}
randomization(N_total = 140, block = 6, allocation = c(2, 1))
```

In the code above, the total sample size is 140, the block size is 6 and the randomization ratio is 2:1 for control to treatment. As $2+1 = 3$ is a multiple of the block size, this allocation is allowed.

Complete randomization can be performed by setting the block size equal to the total sample size as illustrated in the example below. 
```{r}
randomization(N_total = 120, block = 120, allocation = c(2, 1))
```






<!-- # Binomial -->

<!-- ```{r} -->

<!-- value <-  -->
<!--   binomial_outcome(p_treatment_true = 0.08) %>% -->
<!--   historical_binomial(y0_treatment = 5, N0_treatment = 55,  -->
<!--                       y0_control = NULL, N0_control = NULL,  -->
<!--                       discount_function = "identity") %>% -->
<!--   enrollment_rate(lambda = c(0.3, 1), time = 25) %>% -->
<!--   study_details(total_sample_size = 900, study_period = 50,  -->
<!--                 interim_look = c(410, 440, 670), -->
<!--                 prop_loss_to_followup = 0.10, alternative = "less") %>% -->
<!--   impute(no_of_impute = 25, number_mcmc = 1000) %>%  -->
<!--   hypothesis(delta = 0, futility_prob = 0.05, prob_ha = 0.95, expected_success_prob = 0.90) %>% -->
<!--   BACTbinomial(no_of_sim = 110) -->


<!-- str(value) -->

<!-- ``` -->



<!-- # Normal -->

<!-- ```{r} -->
<!-- # value <-  -->
<!-- #   normal_outcome(mu_control_true = 16, mu_treatment_true = 12,  -->
<!-- #                  sd_treatment_true = 1.4, sd_control_true = 1.9) %>% -->
<!-- #   enrollment_rate(lambda = c(0.3, 1), time = 25) %>% -->
<!-- #   study_details(total_sample_size = 300, study_period = 50,  -->
<!-- #                 interim_look = c(210, 240, 270),  -->
<!-- #                 prop_loss_to_followup = 0.10) %>% -->
<!-- #   impute(no_of_impute = 25, number_mcmc = 1000) %>% -->
<!-- #   randomize(block_size = 2, randomization_ratio = c(1, 1)) %>% -->
<!-- #   hypothesis(delta = 0, futility_prob = 0.05, prob_ha = 0.95,  -->
<!-- #              expected_success_prob = 0.90) %>% -->
<!-- #   BACTnormal() -->
<!-- ``` -->

# Historical Data
## Incorporating Historical Data Using Discount Function Approach




