---
title: "bayesCT: An R package for Simulation in Adaptive Bayesian Clinical Trials"
author: "Thevaa Chandereng, Graeme Hickey, Donald Musgrove, Tarek Haddad, Timothy Hanson, Theodore Lystig"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{bayesian trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```


# Introduction

Randomized controlled trials (RCT) are the gold standard of clinical studies of medical and pharmaceutical devices. 
In an RCT, subjects are randomized to an intervention arm, with the standard framework being the comparison of a new treatment/drug to a current treatment/drug (control). 
The majority of research in the design of RCTs and their application has been based on the frequentist paradigm. 
In recent years, the adaptive Bayesian trial design approach has gained attention.

Adaptive Bayesian trials provide much more flexibility over conventional frequentist approaches in terms of the design and analysis of a clinical trial. Notably, adaptive Bayesian trials:

- can incorporate historical data, which helps reduce the sample size required;
- can be exploited for interim analyses of early stopping for success or futility;
- can be used in trials with multiple treatment arms (versus a single control), allowing for treatments not demonstrating success to be dropped; and
- can be used to alter the randomization ratio to improve trial efficiency or even to increase trial recruitment.

A fundamental requirement of an adaptive trial is the *a priori* and interim evaluation of the operating characteristics. In a adaptive trials, these are generally not of closed-form; hence, simulation approaches are required. Simulation of adaptive Bayesian trial designs have historically been substantially more complex compared to frequentist approaches. This package removes this obstacle and allows for simulation of adaptive Bayesian trials under a range of designs and outcome types.

If you use `bayesCT` in published research, please cite: **TBA**.

If after reading through this vignette you have questions or problems using `bayesCT`, please post them to https://github.com/thevaachandereng/bayesCT/issues. This will notify the package maintainers and benefit other users. Please note that comments and questions should **not** be emailed directly to the package authors.

# Running bayesCT

Prior to analyzing your data, the R package needs to be installed.

The easiest way to install `bayesCT` is through CRAN:

```{r, eval = FALSE}
install.packages("bayesCT")
```

There are additional ways to download `bayesCT`. The first option is most useful if want to download a specific version of `bayesCT` (which can be found at https://github.com/thevaachandereng/bayesCT/releases):

```{r, eval = FALSE}
devtools::install_github("thevaachandereng/bayesCT@vx.xx.x")

# or 

devtools::install_version("bayesCT", version = "x.x.x", repos = "http://cran.us.r-project.org")
```

The second option is to download the most recent stable version through GitHub:

```{r, message = FALSE, prompt=FALSE}
devtools::install_github("thevaachandereng/bayesCT")
```

After successful installation, the package must be loaded into the working space:

```{r lib, results="asis", eval=TRUE}
library(bayesCT)
```


# Recruitment

The recruitment of subjects is assumed to follow a Poisson process. 
Since trial recruitment is independent, the ["memoryless property"](https://en.wikipedia.org/wiki/Memorylessness) modeling of subject recruitment. 
However, we cannot assume a homogeneous Poisson process since subject recruitment varies over time. Thus, the `enrollment()` function assumes a piecewise stationary Poisson process with different $\lambda$ parameters over different time intervals. 
The function also omits an enrollment date considering time zero as study initiation. 

For instance, if we have the following piecewise function of $\lambda$, with $t$ representing the time interval,  

\[  \lambda = \left\{
\begin{array}{ll}
      0.3 & t \in [0, 5) \\
      0.7 & t \in [5, 10) \\
      0.9 & t \in [10, 15) \\
      1.2 & t \in [15, \infty) \\
\end{array} 
\right. \]

then to simulate individual patient enrollment dates with a sample size (`N_total`) of 50, one would use

```{r}
enrollment(param = c(0.3, 0.7, 0.9, 1.2), 50, time = c(5, 10, 15))
```


# Randomization scheme

The randomization allocation is important component in the design of an RCT. 
Complete randomization is not always ideal due to the chance of drawing a large block of a single treatment arm, which can impact the time to enrollment completion. 
Therefore, a block randomization allocation is generally preferable. 
Block randomization allocation allows for different randomization ratios but they must be in integer form.
In addition, the block size should also be an integer that is divisible by the sum of randomization allocation. 

For instance,

```{r, eval = FALSE}
randomization(N_total = 140, block = 7, allocation = c(2, 1))
```

will not work because the sum of the allocation is not a multiple of the block. 

```{r}
randomization(N_total = 140, block = 6, allocation = c(2, 1))
```

In the code above, the total sample size is 140, the block size is 6 and the randomization ratio is 2:1 for control to treatment. 
As $2+1 = 3$ is a multiple of the block size, this allocation is allowed.

Complete randomization can be performed by setting the block size equal to the total sample size as illustrated in the example below. 
```{r}
randomization(N_total = 120, block = 120, allocation = c(2, 1))
```






<!-- # Binomial -->

<!-- ```{r} -->

<!-- value <-  -->
<!--   binomial_outcome(p_treatment_true = 0.08) %>% -->
<!--   historical_binomial(y0_treatment = 5, N0_treatment = 55,  -->
<!--                       y0_control = NULL, N0_control = NULL,  -->
<!--                       discount_function = "identity") %>% -->
<!--   enrollment_rate(lambda = c(0.3, 1), time = 25) %>% -->
<!--   study_details(total_sample_size = 900, study_period = 50,  -->
<!--                 interim_look = c(410, 440, 670), -->
<!--                 prop_loss_to_followup = 0.10, alternative = "less") %>% -->
<!--   impute(no_of_impute = 25, number_mcmc = 1000) %>%  -->
<!--   hypothesis(delta = 0, futility_prob = 0.05, prob_ha = 0.95, expected_success_prob = 0.90) %>% -->
<!--   BACTbinomial(no_of_sim = 110) -->


<!-- str(value) -->

<!-- ``` -->



<!-- # Normal -->

<!-- ```{r} -->
<!-- # value <-  -->
<!-- #   normal_outcome(mu_control_true = 16, mu_treatment_true = 12,  -->
<!-- #                  sd_treatment_true = 1.4, sd_control_true = 1.9) %>% -->
<!-- #   enrollment_rate(lambda = c(0.3, 1), time = 25) %>% -->
<!-- #   study_details(total_sample_size = 300, study_period = 50,  -->
<!-- #                 interim_look = c(210, 240, 270),  -->
<!-- #                 prop_loss_to_followup = 0.10) %>% -->
<!-- #   impute(no_of_impute = 25, number_mcmc = 1000) %>% -->
<!-- #   randomize(block_size = 2, randomization_ratio = c(1, 1)) %>% -->
<!-- #   hypothesis(delta = 0, futility_prob = 0.05, prob_ha = 0.95,  -->
<!-- #              expected_success_prob = 0.90) %>% -->
<!-- #   BACTnormal() -->
<!-- ``` -->

# Historical Data

Since bayesian attitude is to bring all the data available to address the scientific question proposed. 
Any particular trial conducted is unlikely to be the first trial conducted in a particular disease.
Pilot studies are conducted in small sample size to obtain estimation of parameters for a clinical trial (clinically relevant effect size, guide sample size determination, recruitment rates). 
Besides pilot study, the control drug/treatment is usually experimented before with other drugs as control/experimental drug before being accepted as a standard drug/treatment to treat a specific disease.
However, statisticians need to careful consider the inclusion and exclusion criterias in a trial before completely accepting a trial data to incorporate it in a new trial as an historical data.
The R package `bayesDP` is used as a computational engine for computing posteriors as well as integrating historical data.



## Incorporating Historical Data Using Discount Function Approach

There are multiple methods to incorporate historical data into a trial.
In our implementation, historical data is integrated using the discount function approach. 
The discount function approach treats historical data with increased skepticism compared to other approaches. 
The limitations of previous dynamic borrowing approaches such as non-negligible weighting of historical data in the presence of substantial disagreement with current data, negatively affecting the type I error, bias and mean squared error (MSE). 
The discount function approach 

In the first estimation step, the historical data weight $\hat{\alpha}$ is estimated. In the case of a two-arm trial, where both treatment and control data are available, $\hat{\alpha}$ value is estimated separately for each of the treatment and control arms. Of course, historical treatment or historical control data must be present, otherwise $\hat{\alpha}$ is not estimated for the corresponding arm.

Details on computation of $\hat\alpha$ is available on the vignette of R package `bayesDP`.





# Stopping Early for Futility or Success

Unlike the conventional clinical trials, adaptive bayesian clinical trials offer much more flexibility. 
The trial can be stopped early for success or futility. 
In our implementation for designing bayesian trials, we compute early success and futility upon imputing loss to follow up. 


## Early Success

For a double arm trial, the early success is specified as the equation below.

\[
P(\underbrace{\theta_T - \theta_C }_{\text{difference in posterior mean}} > \underbrace{\delta}_{\text{margin}}| y, y_0, \alpha) > \underbrace{\Delta}_{\text{Specified success rate}}
\]


If the proprotion of  mean of the treatment difference is greater than a specified margin (default = 0) is greater than the specified success rate, the trial stops for early success. The default specified success rate is set to 0.90 (default). However, the proportion of mean of the treatment difference is not taken as the final posterior estimate since the estimate is obtained from imputation of loss to follow up. 

For a single arm, 

\[
P(\underbrace{\theta_T }_{\text{posterior mean}} > \underbrace{\delta}_{\text{margin}}| y, y_0, \alpha) > \underbrace{\Delta}_{\text{Specified success rate}}
\]

For a single arm, the estimation is done a little differently. More details shown in the simulation portion.  


## Futility

For a double arm trial, the early success is specified as the equation below.

\[
P(\underbrace{\theta_T - \theta_C }_{\text{difference in posterior mean}} > \underbrace{\delta}_{\text{margin}}| y, y_0, \alpha) < \underbrace{\omega}_{\text{Specified futility rate}}
\]


If the proprotion of  mean of the treatment difference is greater than a specified margin (default = 0) is less than the specified futility rate, the trial stops for futility. The default specified success rate is set to 0.10 (default). However, the proportion of mean of the treatment difference is not taken as the final posterior estimate since the estimate is obtained from imputation of loss to follow up. 

For a single arm, 

\[
P(\underbrace{\theta_T }_{\text{posterior mean}} > \underbrace{\delta}_{\text{margin}}| y, y_0, \alpha) < \underbrace{\omega}_{\text{Specified futility rate}}
\]

For a single arm, the estimation is done a little differently. More details shown in the simulation portion.  

