---
title: "Binomial Outcome"
author: "Thevaa Chandereng, Donald Musgrove, Tarek Haddad, Graeme Hickey, Timothy Hanson, Theodore Lystig"
header-includes:
   - \usepackage{amsmath}
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{bayesCT:binomial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
set.seed(28999)
```

```{r, echo = FALSE, message = FALSE}
library(bayesCT)
```


# Introduction

The purpose of this vignette is to introduce how to simulate and analyze an adaptive bayesian clinical trial for binomial outcomes. 
The simulation section compromise of the design of the trial itself which provides a type I error rate and power at each interim look. 
We use the beta-binomial conjugate for the estimation of posterior probabilities. 
The historical data is used as an informative prior and we use `bayesDP` as an engine to incorporate the historical data. 
The default model uses a non-informative prior of $Beta(a = 1, b = 1)$ with/without the incorporation of historical data.
Instead of using traditional R function, we use pipes to input our parameters. 

# Estimation of Treatment Difference 

Let $y$ and $N$ denote the number of events and sample size. 
Let $a_0$ and $b_00$ denote the rate parameters of a Beta distribution.
Then, the posterior distribution of the event rates under vague (flat) priors is

$$ \tilde{\theta}\mid y,N \; \sim \;  \mathcal{B}eta\left(y+a_0,\,N-y+b_0 \right).$$
When historical data is present, then $y_0$ and $N_0$ denote the number of events and sample size of the historical data. 
The weight of the historical data included in the study design and analysis is denoted by $\hat\alpha$. 
For more details on computation of $\hat{\alpha}$, please refer to the vignette of binomial counts avaialable at [https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpbinomial-vignette.html](https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpbinomial-vignette.html).
The posterior distribution of the event rates with historical data incorporated under vague (flat) priors is 

$$\theta \mid y,N, y_0, N_0 \; \sim \; \mathcal{B}eta\left(y+y_0\hat{\alpha}+a_0,\, N-y+\hat{\alpha}(N_0-y_0)+b_0 \right).$$

Since there is no closed-form solution for difference in beta distribution ( $\mathcal{B}eta(\alpha_1, \beta_1)$ - $\mathcal{B}eta(\alpha_2, \beta_2)$), we use Monte Carlo estimation of the posteriors of treatment and control group to compute the difference. 

The estimation of treatment difference $\tilde{\theta_T} - \tilde{\theta_C}$, where $\theta_T$ is the posterior of the treatment group and $\theta_C$ is the posterior of the control group.


# Wrapper Function for Design and Analysis

Unlike traditional R function, the R package depends on pipe inputs with different wrapper functions. All the wrapper functions are illustrated below with details on the arguments for the simulation and analysis. 

**binomial_outcome**: The binomial outcome is a wrapper function for the simulation. The binomial outcome has two input **p_treatment** and **p_control**. 

- **p_treatment** is a required input where the proportion of events for a single-arm or double-arm trials are specified for the treatment arm, where $0 < p_{treatment} < 1$. 

- **p_control** is set to NULL as default for single-arm trials. However, for two-arm trials **p_control** input specifies the proportion of events for the control arm, where $0 < p_{control} < 1$. 


**study_details**: The study details is a wrapper function for the simulation. The **study_details** function takes four input (**total_sample_size**, **study_period**, **interim_look** and **prop_loss_to_followup**). 

- **total_sample_size** input determines the number of sample size required to run the simulation. 

- **study_period** input determines the length of the study. Since interim look at x is done when the x^{th} patient is enrolled, the length of study is important for imputation purpose.  

- **interim_look** function allows a vector of numeric value, where the trials are stopped to check for early success or futility. For more details on the early stopping for futility or success, please look at the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html). The interim_look values need to be smaller than the sample_size value. 

- Loss to follow up is not peculiar in a clinical trial setting, the **prop_loss_to_followup** input allows to set the proportion of loss to follow up. The default value is set to 0.10. The **prop_loss_to_followup** needs to be between 0 and 1.  


**enrollment_rate**: The enrollment rate is a wrapper function for the simulation. The **enrollment_rate** function takes two input (**lambda**, **time**) unlike the **enrollment** function. 

- **lambda** input is similar to **param** input in the **enrollment** function where it takes different lambda values for different poison rates (the default is set to 0.3, i.e. 0.3 patients per recruited per day). 

- **time** input works similarly to the **time** input in the **enrollment** function where **time** input determines the cutoff rate for different poison rate (default set to NULL). For more details, please look at the recruitment section on the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html).  


**hypothesis**: The hypothesis is a wrapper function for both simulation and analysis. The **hypothesis** wrapper takes five different inputs (**delta**, **prob_accept_ha**, **futility_prob**, **expected_success_prob** and **alternative**). 

- The **delta** input determines the threshold difference for treatment and control group or the threshold for single-arm trial (the default is set to 0). For non-inferiority trial, the difference needs to be < 0.  

- The input **prob_accept_ha** controls the probability of accepting the alternative hypothesis. The default is set to 0.95. 

- **futility_prob** input controls the futility rate $\omega$ in the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html). The default is set to 0.10. To prevent the trial for stopping for futility, set the **futility_prob** to 0. 

- **expected_success_prob** input controls the early success rate $\Delta$ in the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html). The default is set to 0.90. To prevent the trial for stopping for early success, set the **expected_success_prob** to 1. For more details on the criteria of early stopping for success or futility, please see the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html). The -

- **alternative** input takes a character string specifying the alternative hypothesis, must be one of "greater" $p_{treatment} > p_{control}$ (default), "less" $p_{treatment} < p_{control}$ or "two-sided" $p_{treatment} \neq p_{control}$.


**randomize**: The randomize is a wrapper function for the simulation. The **randomize**
 function takes two input (**block_size** and **randomization_ratio**). 
 
 - **block_size** input controls the size of each block to randomize. 
 
- **randomization_ratio** is a vector of length 2 which determine the ratio for control to treatment. For more details, please look at the randomization section on the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html).  


**impute**: The impute is a wrapper function for simulation and analysis. The **impute** function takes two input (**no_of_impute**, **number_mcmc**). 

- **no_of_impute** determines the number of imputation done for loss to follow-up to stop for early success and number of imputation done for loss to follow-up and patients not enrolled (only in simulation scenario) for loss to follow-up for futility. The default is set to 10 to generate the vignette quickly. However, the general rule of thumb for implementation of adaptive trials would be to set **no_of_impute** to 10,000. 

- A Monte Carlo estimation approach is used to estimate the posterior probability, requiring several samples from the posterior distributions. The **number_mcmc** controls the number of Monte Carlo estimation with default set to 10000.  


**beta_prior**: The beta prior is a wrapper function for the non-informative prior for simulation and analysis. The **beta_prior** takes two input (**a_0** and **b_0**).

- **a_0** regulates the shape value $\alpha$ in the beta distribution. The default value of **a_0** is 1. To completely avoid using non-informative prior, users can set **a_0** to 0. 

- **b_0** regulates the shape value $\beta$ in the beta distribution. The default value of **b_0** is 1. To completely avoid using non-informative prior, users can set **b_0** to 0. 


**historical_binomial**: The historical binomial wrapper function allows incorporation of historical data using discount function in the analysis and simulation. This wrapper is only used when historical data is incorporated in the analysis or simulation. For more details on the method and computation, please look at the [bayesDP binomial vignette](https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpbinomial-vignette.html). 
Historical binomial takes 9 inputs.

- **y0_treatment**: The number of events in the historical treatment group.

- **N0_treatment**: The sample size for the historical treatment group.

- **y0_control**: The number of events in the historical control group.

- **N0_control**: The sample size for the historical control group.

- **discount_function**: Specify the discount function to use. Currently supports weibull, scaledweibull, and identity. The discount function scaledweibull scales the output of the Weibull CDF to have a max value of 1. The identity discount function uses the posterior probability directly as the discount weight. Default value is "identity".

- **alpha_max**: Maximum weight the discount function can apply. Default is 1. For a two-arm trial, users may specify a vector of two values where the first value is used to weight the historical treatment group and the second value is used to weight the historical control group.

- **fix_alpha**: Fix alpha at alpha_max? Default value is FALSE.

- **weibull_scale**: Scale parameter of the Weibull discount function used to compute alpha, the weight parameter of the historical data. Default value is 0.135. For a two-arm trial, users may specify a vector of two values where the first value is used to estimate the weight of the historical treatment group and the second value is used to estimate the weight of the historical control group. Not used when discount_function = "identity".

- **weibull_shape**: Shape parameter of the Weibull discount function used to compute alpha, the weight parameter of the historical data. Default value is 3. For a two-arm trial, users may specify a vector of two values where the first value is used to estimate the weight of the historical treatment group and the second value is used to estimate the weight of the historical control group. Not used when discount_function = "identity".


**data_binomial**: The data binomial is a wrapper function for inputting data for analysis of bayesian trial with binomial outcome. The **data_binomial** takes three input (**treatment**, **outcome** and **complete**).

- **treatment**: treatment assignment for patients in the trial, 1 for treatment group and 0 for control group. 

- **outcome**: binomial outcome of the trial, 1 for response (success or failure), 0 for no response. The length of the vector is the same size as the treatment variable. 

- **complete**: similar length as treatment and outcome variable, 1 for complete outcome, 0 for loss to follow up. If complete is 0, the outcome of the patient is ignored. If complete is not provided, the dataset is assumed to be complete.


# Design



```{r}
value <-
   binomial_outcome(p_treatment = 0.08) %>%
   study_details(total_sample_size = 900, study_period = 50,
                 interim_look = NULL,
                 prop_loss_to_followup = 0.10)

output <- value %>%
            simulate(no_of_sim = 10)


str(output)
```




```{r}
value <- value %>%
  study_details(total_sample_size = 900, study_period = 50,
                interim_look = c(410, 440, 670),
                prop_loss_to_followup = 0.10)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

```{r}
value <- value %>%
  enrollment_rate(lambda = c(0.3, 1), time = 25)


output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```




```{r}
value <- value %>%
   hypothesis(delta = -0.03, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less")

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```



```{r}
value <- value %>%
  randomize(block_size = c(4, 6), randomization_ratio = c(1, 1))

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```



```{r}
value <- value %>%
  impute(no_of_impute = 25, number_mcmc = 1000)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

```{r}
value <- value %>%
  beta_prior(a0 = 5, b0 = 5)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

```{r}
value <- value %>%
  historical_binomial(y0_treatment = 5, N0_treatment = 55,
                      discount_function = "identity",
                      y0_control = NULL, N0_control = NULL,
                      alpha_max = 1, fix_alpha = FALSE,
                       weibull_scale = 0.135, weibull_shape = 3)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```



```{r}

value <-
  binomial_outcome(p_treatment = 0.08) %>%
  enrollment_rate(lambda = c(0.3, 1), time = 25) %>%
  study_details(total_sample_size = 900, study_period = 50,
                interim_look = c(410, 440, 670),
                prop_loss_to_followup = 0.10) %>%
  hypothesis(delta = -0.03, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  randomize(block_size = c(4, 6), randomization_ratio = c(1, 1)) %>%
  impute(no_of_impute = 25, number_mcmc = 1000) %>%
  beta_prior(a0 = 5, b0 = 5) %>%
  historical_binomial(y0_treatment = 5, N0_treatment = 55,
                      discount_function = "identity",
                      y0_control = NULL, N0_control = NULL,
                      alpha_max = 1, fix_alpha = FALSE,
                       weibull_scale = 0.135, weibull_shape = 3) %>%
  simulate(no_of_sim = 5)


str(value)

```


# Analysis

```{r}
data(binomialdata)

head(binomialdata)
```


```{r}
input <-
  data_binomial(treatment = binomialdata$treatment,
                outcome   = binomialdata$outcome,
                complete  = binomialdata$complete)

out <- input %>%
  analysis(type = "binomial")

str(out)
```



```{r}
input <- input %>%
  hypothesis(delta = 0.02, futility_prob = 0.05, prob_accept_ha = 0.95,                                       expected_success_prob = 0.90, alternative = "greater")

out <- input %>%
  analysis(type = "binomial")

str(out)
```



```{r}

input <- input %>%
  impute(no_of_impute = 50, number_mcmc = 10000)

out <- input %>%
  analysis(type = "binomial")

str(out)
```


```{r}

input <- input %>%
  beta_prior(a0 = 3, b0 = 3)

out <- input %>%
  analysis(type = "binomial")

str(out)

```


```{r}
input <- input %>%
  historical_binomial(y0_treatment = 5, N0_treatment = 55,
                      y0_control = NULL, N0_control = NULL,
                      discount_function = "weibull",
                      alpha_max = 1, fix_alpha = FALSE,
                      weibull_scale = 0.135, weibull_shape = 3)

out <- input %>%
  analysis(type = "binomial")

str(out)

```



```{r}
ouut <-
  data_binomial(treatment = binomialdata$treatment,
                outcome   = binomialdata$outcome,
                complete  = binomialdata$complete) %>%
  hypothesis(delta = 0.02, futility_prob = 0.05, prob_accept_ha = 0.95,                                       expected_success_prob = 0.90, alternative = "greater") %>%
  impute(no_of_impute = 50, number_mcmc = 10000) %>%
  beta_prior(a0 = 5, b0 = 5) %>%
  historical_binomial(y0_treatment = 5, N0_treatment = 55,
                      y0_control = NULL, N0_control = NULL,
                      discount_function = "identity",
                      alpha_max = 1, fix_alpha = FALSE,
                      weibull_scale = 0.135, weibull_shape = 3) %>%
  analysis(type = "binomial")

str(out)

```

