---
title: "Continuous Outcomes"
author: "Thevaa Chandereng, Donald Musgrove, Graeme L. Hickey"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, binomial]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Continuous Outcomes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
suppressWarnings(RNGversion("3.5.0"))
set.seed(28999)
```

```{r, message=FALSE, echo=FALSE}
library(bayesCT)
```

# Introduction

The purpose of this vignette is to introduce how to simulate and analyze an adaptive Bayesian clinical trial for continuous-valued, i.e., normal, outcomes. The simulation section compromises the design of the trial itself which provides type I error rate and power at each interim look. We use the normal conjugate prior for the estimation of posterior values. Available historical data can be used as an informative prior; we use the `bayesDP` package as the engine for incorporating the historical data. Instead of using traditional R function, we use pipes to input our parameters. 

# Estimation of treatment difference

Let $\bar{y}$, $s$, and $N$ denote the sample mean, sample standard deviation, and sample size, respectively. Then, the posterior distribution of the mean under vague (flat) priors is

\[
\begin{array}{rcl}
  \tilde{\sigma}^2\mid\bar{y},s,N & \sim & InverseGamma\left(\frac{N-1}{2},\,\frac{N-1}{2}s^2 \right),\\
  \\
  \tilde{\mu}\mid\bar{y},N,\tilde{\sigma}^2 & \sim & \mathcal{N}ormal\left(\bar{y},\, \frac{1}{N}\tilde{\sigma}^2  \right).
\end{array}
\]

When historical data is present, let $\bar{y}_0$, $s_0$, and $N_0$ denote the sample mean, sample standard deviation, and sample size of the historical data, respectively. The posterior distribution of the mean for the historical data is

\[
\begin{array}{rcl}
  \sigma^2_0\mid\bar{y_0},s_0,N_0 & \sim & InverseGamma\left(\frac{N_0-1}{2},\,\frac{N_0-1}{2}s_0^2 \right),\\
  \\
  \mu_0 \mid \bar{y}_0, N_0, \sigma^2_0 & \sim & \mathcal{N}ormal\left(\bar{y}_0,\,\frac{1}{N_0}\sigma^2_0 \right).
\end{array}
\]

The weight of the historical data included in the study design and analysis is denoted by $\hat{\alpha}$. 
For more details on computation of $\hat{\alpha}$, please refer to the vignette of binomial counts available at [https://CRAN.R-project.org/package=bayesDP](https://CRAN.R-project.org/package=bayesDP).
The posterior distribution of the mean outcome with historical data incorporated under vague (flat) prior distributions is:

\[
\tilde{\mu} \sim \mathcal{N}ormal\left( \frac{\sigma^2_0N\bar{y} + \tilde{\sigma}^2N_0\bar{y}_0\hat{\alpha}}{N\sigma^2_0 + \tilde{\sigma}^2N_0\hat{\alpha}},\,\frac{\tilde{\sigma}^2\sigma^2_0}{N\sigma^2_0 + \tilde{\sigma}^2N_0\hat{\alpha}} \right).
\]

Even though there is a closed-form solution for the difference in normally distributed random variables, we use Monte Carlo simulations to estimate the treatment difference. 

The estimation of the treatment difference is $\tilde{\mu_T} - \tilde{\mu_C}$, where $\mu_T$ is the posterior mean outcome in the treatment group and $\mu_C$ is the posterior mean outcome in the control group.

# Wrapper function for design and analysis

Unlike traditional R functions, the `bayesCT` package depends on pipe inputs with different wrapper functions. The key wrapper functions are:

* `normal_outcome`: specifying means and standard deviations.
* `study_details`: specifying sample size, study length, interim looks, and loss to follow-up.
* `enrollment_rate`: specifying enrollment.
* `hypothesis`: specifying the hypothesis for both simulation and analysis.
* `randomize`: specifying the randomization scheme. 
* `impute`: specifying imputation of outcomes at the interim looks.
* `beta_prior`: specifying the non-informative prior. 
* `historical_binomial`: specifying historical data, if available. This function should only be used when historical data is incorporated in the analysis or simulation. For more details on the method and computation, please see [https://CRAN.R-project.org/package=bayesDP](https://CRAN.R-project.org/package=bayesDP).
* `data_normal`: inputting data with a normally distributed outcome. 
* `analysis`: analyzing a trial at an interim analysis.
* `simulate`: simulating trials. **Note:** should only be used as the terminal end of the pipe. 

# Design of adaptive trials

In the following section, we will discuss the design of adaptive trials using `bayesCT` for normal outcomes. We illustrate an example for one-arm trial and two-arm trials using the wrapper functions described above. 

## One-arm trial

In the example below, we will illustrate how to compute power, type 1 error, and other clinical trial characteristics for an objective performance criterion (OPC) (or performance goal; PG) trial with mean outcome and hypothesis described as follows,

\[
H_0: \mu_{treatment} \geq 120 \qquad H_A:\mu_{treatment} < 120.
\]

The most important wrapper functions are `study_details` and `normal_outcome` (especially since there are no default values).

The normally distributed outcomes are simulated using a mean value of 120 and standard deviation of 5.5. The total sample size is 400 with a study length of 60 days. A 10% loss to follow up is assumed. Based on this information, the adaptive trials are simulated 2 times to obtain the following output (**Note**: for the purpose of generating the vignette quickly, we reduce the number of simulations to 5, you should use a much larger value, e.g., 10,000). The aforementioned inputs were chosen for illustration purposes only.

```{r opcminimum}
value <- normal_outcome(mu_treatment = 120,
                        sd_treatment = 5.5) %>%
  study_details(total_sample_size     = 400, 
                study_period          = 60,
                interim_look          = NULL,
                prop_loss_to_followup = 0.10)
				
# Simulate 2 trials
output <- value %>%
  simulate(no_of_sim = 2)

# Structure of the simulation output
str(output)
```

To allow for early stopping for success or futility, we add interim looks to the design. We'll check for success or futility at the enrollment of the 350th and 380th subject. Upon adding this interim look requirement, the trial is simulated 2 times to obtain the output. 

```{r opcinterimlook}
# adding interim look
value <- value %>%
  study_details(total_sample_size     = 400, 
                study_period          = 60,
                interim_look          = c(350, 380),
                prop_loss_to_followup = 0.10)

# Simulate 2 trials
output <- value %>%
  simulate(no_of_sim = 2)

str(output)
```

Patient enrollment is assumed to follow a Poisson process. The default enrollment rate is 0.3 patients per day. In this simulation we'll introduce a step-wise Poisson process with rate $\lambda$ as follows:

\[
\lambda = \left\{
\begin{array}{ll}
      0.4 & \text(time) \in [0, 40) \\
      0.7 & \text(time) \in [40, \infty) \\
\end{array} 
\right.
\]

This enrollment scheme is illustrated below. 

```{r opcenroll}
value <- value %>%
  enrollment_rate(lambda = c(0.4, 0.7), 
                  time = 40) 

output <- value %>%
  simulate(no_of_sim = 2)
```

The hypothesis is an important wrapper function which controls the probability of futility, probability of accepting the alternative hypothesis, probability of early success, the alternative hypothesis, and the treatment difference margin.
 
Since, in an OPC trial, the outcomes in the treatment group are simulated using the input provided, delta controls the maximum threshold allowed for trial to succeed/fail. The default value of `delta` is 0. Here, we'll use `delta = -10` (i.e $120 - \hat{\mu}_{treatment} < -10$). 

We'll further set the futility probability to 0.10, the expected success probability for early stopping to 0.85, and the final probability of accepting the alternative to 0.95. The alternative is `"less"` due to the hypothesis function specified above.

```{r opchypo}
value <- value %>%
   hypothesis(delta                 = -10, 
              futility_prob         = 0.10, 
			        prob_accept_ha        = 0.95,
              expected_success_prob = 0.85, 
			        alternative           = "less")

output <- value %>%
  simulate(no_of_sim = 2)
```

Next, we'll illustrate imputations for imputing outcomes for subjects loss to follow up. We'll carry out 20 imputations and draw 2000 values from the posterior of each imputation. 

```{r opcimpute}
value <- value %>%
  impute(no_of_impute = 20, 
         number_mcmc  = 2000)

output <- value %>%
  simulate(no_of_sim = 2)
```

The above flow was for illustrative purposes. Instead of inputting parameters step by step, the trial parameters can be filled in all at once as illustrated below. The pipe function connects all inputs together and the trial is simulated 2 times to obtain results. 

```{r opcoverall}
value <- normal_outcome(mu_treatment = 120,
                        sd_treatment = 5.5) %>%
  study_details(total_sample_size     = 400, 
                study_period          = 60,
                interim_look          = c(350, 380),
                prop_loss_to_followup = 0.10) %>%
  hypothesis(delta                 = -10, 
             futility_prob         = 0.10, 
			       prob_accept_ha        = 0.95,
             expected_success_prob = 0.85, 
			       alternative           = "less") %>%
  enrollment_rate(lambda = c(0.4, 0.7), 
                  time   = 4) %>%
  randomize(block_size          = c(10, 20), 
            randomization_ratio = c(1, 1)) %>%
  impute(no_of_impute = 5, 
         number_mcmc  = 2000)  %>%
  simulate(no_of_sim = 2)

str(value)
```

## Two-arm trial

In this section, we will illustrate how to perform the design of a two-arm trial with the incorporation of historical data. The example will compute the type 1 error, power, and other outputs for a superiority trial. The study hypothesis is

\[
H_0: \mu_{treatment} - \mu_{control} \leq 0 \qquad H_A: \mu_{treatment} - \mu_{control} > 0.
\]

Unlike the OPC trial above, we will not include interim looks. The normal mean outcomes are simulated using a mean value of 13 and standard deviation of 1.4 for the treatment group and a mean value of 15 and standard deviation of 1.9 for the control group. The total sample size is 400 with a study length of 30 days. A 15% loss to follow up is assumed. The following code simulates a trial 2 times using the piping procedure.

```{r twoarmall}
value <- normal_outcome(mu_treatment = 13, 
                        mu_control = 16, 
						sd_treatment = 1.4, 
						sd_control = 1.9) %>%
  study_details(total_sample_size     = 300, 
                study_period          = 50,
                interim_look          = NULL,
                prop_loss_to_followup = 0.10) %>%
  hypothesis(delta                 = 0, 
             futility_prob         = 0, 
			       prob_accept_ha        = 0.95,
             expected_success_prob = 1, 
			       alternative           = "less") %>%
  impute(no_of_impute = 25, 
         number_mcmc  = 5000) %>%
  enrollment_rate(lambda = c(0.8), 
                  time = NULL) %>%
  randomize(block_size          = c(4, 6), 
            randomization_ratio = c(1, 1)) %>%
  historical_normal(mu0_treatment     = 13, 
                    sd0_treatment     = 5, 
					          N0_treatment      = 100,
                    mu0_control       = 12, 
					          sd0_control       = 3, 
					          N0_control        = 120, 
                    discount_function = "scaledweibull", 
                    alpha_max         = FALSE, 
					          fix_alpha         = 1,
                    weibull_scale     = 0.135, 
					          weibull_shape     = 3,
                    method            = "fixed") %>%
  simulate(no_of_sim = 2)

str(value)
```

# Analysis

In this section, we will demonstrate how to run an adaptive Bayesian trial using `bayesCT`. A sample dataset is provided in the package. The dataset `normaldata` contains the results of 300 subjects from a two-arm trial with normal outcome. The `complete` column indicates whether the outcome was observed, i.e. loss to follow-up.

```{r data}
data(normaldata)
head(normaldata)
```

The minimum input needed to run an adaptive Bayesian trial is the data itself. The `data_normal` input allows the input of the data. The treatment group (0 for control, 1 for treatment) and outcome input are essential for the analysis. However, if the complete input is not provided, the function assumes the outcome data is complete. A default analysis is carried out below. 

```{r analysisdatainput}
input <- data_normal(treatment = normaldata$treatment, 
                     outcome   = normaldata$outcome, 
                     complete  = normaldata$complete) 

out <- input %>%
  analysis(type = "normal")
```

We'll now illustrate using piping to carry out the complete analysis. First, we'll assume the following hypothesis:

\[
H_0:\mu_{treatment} - \mu_{control} > 0 \quad H_A: \mu_{treatment} - \mu_{control} \le 0
\]

The delta and alternative used to analyze the trial is 0 and `"less"` respectively. The probability of accepting the alternative is 0.95, the probability of stopping for futility is 0.05, and the probability of stopping for success is 0.90. We will carry out imputations on subjects loss to follow up. Additionally, we will incorporate historical data on the treatment arm.

```{r analysisall}
out <- data_normal(treatment = normaldata$treatment,
                   outcome   = normaldata$outcome, 
                   complete  = normaldata$complete) %>%
  hypothesis(delta                 = 0, 
             futility_prob         = 0.05, 
			       prob_accept_ha        = 0.95,
             expected_success_prob = 0.90, 
			       alternative           = "less") %>%
  impute(no_of_impute = 10, 
         number_mcmc  = 8000) %>%
  analysis(type = "normal")

str(out)
```
