---
title: "Normal Outcome"
author: "Thevaa Chandereng, Donald Musgrove, Tarek Haddad, Graeme Hickey, Timothy Hanson, Theodore Lystig"
header-includes:
   - \usepackage{amsmath}
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{bayesCT:normal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
set.seed(28999)
```

```{r, message = FALSE, echo = FALSE}
library(bayesCT)
```

# Introduction

The purpose of this vignette is to introduce how to simulate and analyze an adaptive bayesian clinical trial for normal outcomes. 
The simulation section compromise of the design of the trial itself which provides a type I error rate and power at each interim look. 
We use the normal-normal conjugate for the estimation of posterior probabilities. 
The historical data is used as an informative prior and we use `bayesDP` as an engine to incorporate the historical data. 


# Estimation of Treatment Difference
Let $\bar{y}$, $s$, and $N$ denote the sample mean, sample standard deviation, and sample size, respectively. 
Then, the posterior distribution of the mean under vague (flat) priors is

$$ \begin{array}{rcl}
\tilde{\sigma}^2\mid\bar{y},s,N & \sim & InverseGamma\left(\frac{N-1}{2},\,\frac{N-1}{2}s^2 \right),\\
\\
\tilde{\mu}\mid\bar{y},N,\tilde{\sigma}^2 & \sim & \mathcal{N}ormal\left(\bar{y},\, \frac{1}{N}\tilde{\sigma}^2  \right).
\end{array}$$

When historical data is present, let $\bar{y}_0$, $s_0$, and $N_0$ denote the sample mean, sample standard deviation, and sample size of the historical data, respectively.
The posterior distribution of the mean for the historical data is 
$$ \begin{array}{rcl}
\sigma^2_0\mid\bar{y_0},s_0,N_0 & \sim & InverseGamma\left(\frac{N_0-1}{2},\,\frac{N_0-1}{2}s_0^2 \right),\\
\\
\mu_0 \mid \bar{y}_0, N_0, \sigma^2_0 &  \sim  & \mathcal{N}ormal\left(\bar{y}_0,\, \frac{1}{N_0}\sigma^2_0  \right).
\end{array}$$
The weight of the historical data included in the study design and analysis is denoted by $\hat\alpha$. 
For more details on computation of $\hat{\alpha}$, please refer to the vignette of binomial counts avaialable at [https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpnormal-vignette.html](https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpnormal-vignette.html).
The posterior distribution of the mean outcome with historical data incorporated under vague (flat) priors is 
$$\tilde{\mu} \sim \mathcal{N}ormal\left( \frac{\sigma^2_0N\bar{y} + \tilde{\sigma}^2N_0\bar{y}_0\hat{\alpha}}{N\sigma^2_0 + \tilde{\sigma}^2N_0\hat{\alpha}},\,\frac{\tilde{\sigma}^2\sigma^2_0}{N\sigma^2_0 + \tilde{\sigma}^2N_0\hat{\alpha}}      \right).$$

Even though there is a closed-form solution for difference in normal distribution ( $\mathcal{N}ormal(\mu_1, \sigma_1^2)$ - $\mathcal{N}ormal(\mu_2, \sigma_2^2)$), we use Monte Carlo estimation of the posteriors of treatment and control group to compute the effect of treatment difference. 

The estimation of mean of treatment difference $\tilde{\mu_T} - \tilde{\mu_C}$, where $\mu_T$ is the posterior mean outcome in the treatment group and $\mu_C$ is the posterior mean outcome in the control group.




# Wrapper Function for Design and Analysis

Unlike traditional R function, the R package depends on pipe inputs with different wrapper functions. All the wrapper functions are illustrated below with details on the arguments for the simulation and analysis. 

**normal_outcome**: The normal outcome is a wrapper function for the simulation. The normal outcome has four input (**mu_treatment**, **mu_control**, **sd_treatment**, **sd_control**).

- **mu_treatment** and **sd_treatment** are required inputs where the mean and standard deviation outcome for a single-arm or double-arm trials are specified for the treatment arm, where $ sd_{treatment} > 0$. 

- **mu_control** and **sd_control** are set to NULL as default for single-arm trials. However, for two-arm trials **mu_control** and **sd_control** inputs specify the mean and standard deviation outcome for the control arm, where $sd_{control} > 0$. 


**study_details**: The study details is a wrapper function for the simulation. The **study_details** function takes four input (**total_sample_size**, **study_period**, **interim_look** and **prop_loss_to_followup**). 

- **total_sample_size** input determines the number of sample size required to run the simulation. 

- **study_period** input determines the length of the study. Since interim look at x is done when the x^{th} patient is enrolled, the length of study is important for imputation purpose.  

- **interim_look** function allows a vector of numeric value, where the trials are stopped to check for early success or futility. For more details on the early stopping for futility or success, please look at the [https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#early-stopping-for-futility-or-success](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#early-stopping-for-futility-or-success). The interim_look values need to be smaller than the sample_size value. 

- Loss to follow up is not peculiar in a clinical trial setting, the **prop_loss_to_followup** input allows to set the proportion of loss to follow up. The default value is set to 0.10. The **prop_loss_to_followup** needs to be between 0 and 1.  


**enrollment_rate**: The enrollment rate is a wrapper function for the simulation. The **enrollment_rate** function takes two input (**lambda**, **time**) unlike the **enrollment** function. 

- **lambda** input is similar to **param** input in the **enrollment** function where it takes different lambda values for different poison rates (the default is set to 0.3, i.e. 0.3 patients per recruited per day). 

- **time** input works similarly to the **time** input in the **enrollment** function where **time** input determines the cutoff rate for different poison rate (default set to NULL). For more details, please look at the recruitment section on the [https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#recruitment](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#recruitment).  


**hypothesis**: The hypothesis is a wrapper function for both simulation and analysis. The **hypothesis** wrapper takes five different inputs (**delta**, **prob_accept_ha**, **futility_prob**, **expected_success_prob** and **alternative**). 

- The **delta** input determines the threshold difference for treatment and control group or the threshold for single-arm trial (the default is set to 0). For non-inferiority trial, the difference needs to be < 0.  

- The input **prob_accept_ha** controls the probability of accepting the alternative hypothesis. The default is set to 0.95. 

- **futility_prob** input controls the futility rate $\omega$ in the [https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#futility](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#futility). The default is set to 0.10. To prevent the trial for stopping for futility, set the **futility_prob** to 0. 

- **expected_success_prob** input controls the early success rate $\Delta$ in the [bayesCT main vignette](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html). The default is set to 0.90. To prevent the trial for stopping for early success, set the **expected_success_prob** to 1. For more details on the criteria of early stopping for success or futility, please see the [https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#success](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#success). The -

- **alternative** input takes a character string specifying the alternative hypothesis, must be one of "greater" $mu_{treatment} > mu_{control}$ (default), "less" $mu_{treatment} < mu_{control}$ or "two-sided" $mu_{treatment} \neq mu_{control}$.


**randomize**: The randomize is a wrapper function for the simulation. The **randomize**
 function takes two input (**block_size** and **randomization_ratio**). 
 
 - **block_size** input controls the size of each block to randomize. 
 
- **randomization_ratio** is a vector of length 2 which determine the ratio for control to treatment. For more details, please look at the randomization section on the [https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#randomization-scheme](https://thevaachandereng.github.io/bayesCT/articles/bayesCT.html#randomization-scheme).  


**impute**: The impute is a wrapper function for simulation and analysis. The **impute** function takes two input (**no_of_impute**, **number_mcmc**). 

- **no_of_impute** determines the number of imputation done for loss to follow-up to stop for early success and number of imputation done for loss to follow-up and patients not enrolled (only in simulation scenario) for loss to follow-up for futility. The default is set to 10 to generate the vignette quickly. However, the general rule of thumb for implementation of adaptive trials would be to set **no_of_impute** to 10,000. 

- A Monte Carlo estimation approach is used to estimate the posterior probability, requiring several samples from the posterior distributions. The **number_mcmc** controls the number of Monte Carlo estimation with default set to 10000.  


**historical_normal**: The historical normal wrapper function allows incorporation of historical data using discount function in the analysis and simulation. This wrapper is only used when historical data is incorporated in the analysis or simulation. For more details on the method and computation, please look at the [bayesDP normal vignette](https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpnormal-vignette.html). 
Historical binomial takes 9 inputs.

- **mu0_treatment**: Mean of the historical treatment group.

- **sd0_treatment**: The standard deviation of the historical treatment group.

- **N0_treatment**: The sample size for the historical treatment group.

- **mu0_control**: Mean of the historical control group.

- **sd0_control**: The standard deviation of the historical control group.

- **N0_control**: The sample size for the historical control group.

- **discount_function**: Specify the discount function to use. Currently supports weibull, scaledweibull, and identity. The discount function scaledweibull scales the output of the Weibull CDF to have a max value of 1. The identity discount function uses the posterior probability directly as the discount weight. Default value is "identity".

- **alpha_max**: Maximum weight the discount function can apply. Default is 1. For a two-arm trial, users may specify a vector of two values where the first value is used to weight the historical treatment group and the second value is used to weight the historical control group.

- **fix_alpha**: Fix alpha at alpha_max? Default value is FALSE.

- **weibull_scale**: Scale parameter of the Weibull discount function used to compute alpha, the weight parameter of the historical data. Default value is 0.135. For a two-arm trial, users may specify a vector of two values where the first value is used to estimate the weight of the historical treatment group and the second value is used to estimate the weight of the historical control group. Not used when discount_function = "identity".

- **weibull_shape**: Shape parameter of the Weibull discount function used to compute alpha, the weight parameter of the historical data. Default value is 3. For a two-arm trial, users may specify a vector of two values where the first value is used to estimate the weight of the historical treatment group and the second value is used to estimate the weight of the historical control group. Not used when discount_function = "identity".


**data_normal**: The data normal is a wrapper function for inputting data for analysis of bayesian trial with normal outcome. The **data_normal** takes three input (**treatment**, **outcome** and **complete**).

- **treatment**: treatment assignment for patients in the trial, 1 for treatment group and 0 for control group. 

- **outcome**: normal outcome of the trial. The length of the vector is the same size as the treatment variable. 

- **complete**: similar length as treatment and outcome variable, 1 for complete outcome, 0 for loss to follow up. If complete is 0, the outcome of the patient is ignored. If complete is not provided, the dataset is assumed to be complete.

**analysis**: The analysis is a wrapper function for analyzing of bayesian trial with normal outcome. The **analysis** takes the pipe function input and analyze the trial. 

- **input**: The input of the analysis is a list of input to run the adaptive trials. 

- **type**: The type input determines what trial to be analyzed. In the normal trials, the type needs to be set to "normal". 


**simulate**: The simulate is a wrapper function for simulation of adaptive trials. The **simulate** takes the pipe function input and runs the trial. 

- **input** - The input of the analysis is a list of input to run the adaptive trials. 

- **no_of_sim**- The no_of_sim determines the number of trials to be run. The default is set to 10,000. 

# Design

We will discuss the design of adaptive trial using bayesCT for normal outcome in the following section. 
We illustrate an example for one-arm trial and two-arm trial using the wrapper functions described above. 

## One-arm Trial

In the example below, we will illustrate how to compute power, type 1 error and other clinical trial details for an objective performace criterion (OPC) trial with mean outcome and hypothesis described as follows, 
$$H_0: \mu_{treatment} \geq 120 \qquad H_A:\mu_{treatment} < 120.$$
There is no historical data incorporated in this trial. 
**study_details** and **binomial_outcome** are the most important wrapper function for the design of adaptive trials using bayesCT (since there is no default set).
The normal mean outcomes are simulated using a mean value of 120 and standard deviation of 5.5. 
The total sample size is 400 with a study length of 60 days. 
A 10% loss to follow up are induced to the simulated data at random.
Based on this input, the adaptive trials are simulated 10 times to obtain the following output. 
For the purpose of reproducing the vignette quickly, we reduce the number of simulation to 10. 
The general rule of thumb for number of simulation used in clinical trials is 10,000.
**Note: The inputs used are picked for illustration purpose.**

```{r opcminimum}
value <-
  normal_outcome(mu_treatment = 120,
                 sd_treatment = 5.5) %>%
  study_details(total_sample_size = 400, study_period = 60,
                interim_look = NULL,
                prop_loss_to_followup = 0.10)
output <- value %>%
            simulate(no_of_sim = 10)
str(output)
```

To allow for early stopping for success or futility, we add interim look to the design where the trial is stopped at the enrollment of 350th and 380th patient. 
Upon adding this interim look requirement, the trial is simulated 10 times to obtain the output.


```{r opcinterimlook}
# adding interim look
value <- value %>%
  study_details(total_sample_size = 400, study_period = 60,
                interim_look = c(350, 380),
                prop_loss_to_followup = 0.10)

# simulating 10 times
output <- value %>%
            simulate(no_of_sim = 10)

# structure of the simulation output
str(output)
```

The enrollment of patients in the trial is assumed to be a Poisson process.
The default enrollment rate is assumed to enroll 0.3 patients per day. 
However, in this simulation we introduce a step-wise Poisson process with lambda as follows
\[  \lambda = \left\{
\begin{array}{ll}
      0.4 & t \in [0, 40) \\
      0.7 & t \in [40, \infty) \\
\end{array} 
\right. \]

The following enrollment rate is added to the simulation and the trial is simulated for 10 times.


```{r opcenroll}
value <- value %>%
  enrollment_rate(lambda = c(0.4, 0.7), time = 40) 


output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```



The hypothesis is an important wrapper function which controls the probability of futility, probability of accepting alternative hypothesis, probablity of early success, alternative hypothesis and margin of treatment difference. 
Since, in a OPC trial, the proportion of events in the treatment group are simulated using the input provided, the delta controls the maximum threshold allowed for trial to succeed/fail. 
The default delta used is 0, we set the delta to -10 (i.e $120 - \hat{\mu}_{treatment} < -10$). 
The futility probability is set at 0.10 and the expected success probability for early stopping is set at 0.85 with the final probability of accepting the alternative set at 0.95. 
The alternative is set to "less" due to the hypothesis function above.  
The trial is simulated again 10 times to obtain the output. 


```{r opchypo}
value <- value %>%
   hypothesis(delta = -10, futility_prob = 0.10, prob_accept_ha = 0.95,
             expected_success_prob = 0.85, alternative = "less")

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```


The simulation is done for 20 times to impute the loss to follow up data and the posterior are sampled for 2000 (number_mcmc) to obtain an estimate for the posterior value.
The trial is simulated for 10 times to obtain the output. 

```{r opcimpute}
value <- value %>%
  impute(no_of_impute = 20, number_mcmc = 2000)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

Instead of inputting parameters step by step, the trial parameters can be filled at once as illustrtaed below. 
The pipe function connects all input together and finally the simulation is done for 10 times to obtain the results. 


```{r opcoverall}
value <-
  normal_outcome(mu_treatment = 120,
                 sd_treatment = 5.5) %>%
  study_details(total_sample_size = 400, study_period = 60,
                interim_look = c(350, 380),
                prop_loss_to_followup = 0.10) %>%
  hypothesis(delta = -7, futility_prob = 0.10, prob_accept_ha = 0.95,
             expected_success_prob = 0.85, alternative = "less") %>%
  enrollment_rate(lambda = c(0.4, 0.7), time = 4) %>%
  randomize(block_size = c(10, 20), randomization_ratio = c(1, 1)) %>%
  impute(no_of_impute = 20, number_mcmc = 2000)


```



## Two-arm Trial

In this section, we will illustrate how to perform the design of a two-arm trial with incorporating historical data. 
The example will compute the type 1, power and other outputs for a superiority trial with hypothesis as indicated below, 
$$H_0: \mu_{treatment} - \mu_{control} \leq 0 \qquad H_A: \mu_{treatment} - \mu_{control} > 0.$$
Unlike the previous OPC trial, we do not include interim look in this trial. 
The normal mean outcome are simulated using a mean value of 13 and standard deviation of 1.4 for the treatment group and a mean value of 15 and standard deviation of 1.9 for the control group. 
The total sample size is 400 with a study length of 30 days. 
A 15% loss to follow up are induced to the simulated data at random.
Based on this input, the adaptive trials are simulated 10 times to obtain the following output. 
We simulated the trials for 10 times and the input and ouput are included below.


```{r twoarmmin}
value <-
  normal_outcome(mu_treatment = 13, mu_control = 16,
                 sd_treatment = 1.4, sd_control = 1.9) %>%
  study_details(total_sample_size = 300, study_period = 50,
                interim_look = NULL,
                prop_loss_to_followup = 0.10)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```
The hypothesis is an important wrapper function which controls the probability of futility, probability of accepting alternative hypothesis, probablity of early success, alternative hypothesis and margin of treatment difference. 
The margin of treatment difference $\delta$ is set to 0 since we are designing a superiority trial. 
The futility probability is set at 0 and the expected success probability for early stopping is set at 1 with the final probability of accepting the alternative set at 0.975. 
Since, there is no interim look, the probability of futility and expected success does not affect the trial. 
The alternative is set to "less" due to the alternative hypothesis function above.  
The trial is simulated again 10 times to obtain the output. 

```{r twoarmhypo}
value <- value %>%
  hypothesis(delta = 0, futility_prob = 0, prob_accept_ha = 0.95,
             expected_success_prob = 1, alternative = "less")
  

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

The enrollment of patients in the trial is assumed to be a Poisson process.
The enrollment rate in this trial is assumed to be 0.8 patients per day. 
To induce variation, we simulated the randomization ratio using block size of 4 and 6 repetitively and 1:1 for control to treatment. 
The following randomization scheme is implemented below using randomize wrapper and the trial is simulated again 10 times to obtain the output.


```{r twoarmenroll}
value <- value %>%
  enrollment_rate(lambda = c(0.8), time = NULL) %>%
  randomize(block_size = c(4, 6), randomization_ratio = c(1, 1))
  

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```
The simulation is done for 25 times to impute the loss to follow up data and the posterior are sampled for 5000 (number_mcmc) to obtain an estimate for the posterior value.
The trial is simulated for 10 times to obtain the output. 

```{r twoarmimpute}
value <- value %>%
  impute(no_of_impute = 25, number_mcmc = 5000)
  

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```

However, if historical data is avaialble, historical data can be augmented in the analysis using the discount prior approach as implemented in **bayesDP**. 
In our OPC trial, we augmented historical data for illustration purpose. 
The historical data is incorporated for the treatment group using the discount function.
The historical data has a mean of 13, standard deviation of 5 and sample size of 100 for the treatment group.
The historical data has a mean of 12, standard deviation of 3 and sample size of 120 for the control group.
For more details on the method and computation, please look at the [bayesDP normal vignette](https://cran.r-project.org/web/packages/bayesDP/vignettes/bdpnormal-vignette.html). 
**historical_normal** takes 9 inputs as indicated above.
The trial is simulated for 10 times to obtain the output. 


```{r twoarmhistorical}
value <- value %>%
  historical_normal(mu0_treatment = 13, sd0_treatment = 5, N0_treatment = 100,
                    mu0_control = 12, sd0_control = 3, N0_control = 120, 
                    discount_function = "scaledweibull", 
                    alpha_max = F, fix_alpha = 1,
                    weibull_scale = 0.135, weibull_shape = 3)

output <- value %>%
            simulate(no_of_sim = 10)

str(output)
```


Instead of inputting parameters step by step, the trial parameters can be filled at once as illustrtaed below. 
The pipe function connects all input together and finally the simulation is done for 10 times to obtain the results. 


```{r twoarmall}
value <-
  normal_outcome(mu_treatment = 13, mu_control = 16,
                 sd_treatment = 1.4, sd_control = 1.9) %>%
  study_details(total_sample_size = 300, study_period = 50,
                interim_look = NULL,
                prop_loss_to_followup = 0.10) %>%
  hypothesis(delta = 0, futility_prob = 0, prob_accept_ha = 0.95,
             expected_success_prob = 1, alternative = "less") %>%
  impute(no_of_impute = 25, number_mcmc = 5000) %>%
  enrollment_rate(lambda = c(0.8), time = NULL) %>%
  randomize(block_size = c(4, 6), randomization_ratio = c(1, 1)) %>%
  historical_normal(mu0_treatment = 13, sd0_treatment = 5, N0_treatment = 100,
                    mu0_control = 12, sd0_control = 3, N0_control = 120, 
                    discount_function = "scaledweibull", 
                    alpha_max = F, fix_alpha = 1,
                    weibull_scale = 0.135, weibull_shape = 3) %>%
  simulate(no_of_sim = 10)


str(value)
```


# Analysis

In this section, we will demonstrate how to run adaptive bayesian trial using **bayesCT**. 
A sample dataset is provided in the package. 
The dataset **normaldata** contains the results of 300 patients with binomial outcome, treatment group and loss to follow up.

```{r data}
data(normaldata)

head(normaldata)
```

The minimum input needed to run an adaptive bayesian trial is the data itself. 
The **data_normal** input allows the input of the data. 
The treatment group (0 for control, 1 for treatment) and outcome input are essential for the analysis. 
However, if the complete input is not provided, the function assumes the outcome data is complete. 
The data is analyzed using the input below. 

```{r analysisdatainput}
input <-
  data_normal(treatment = normaldata$treatment, 
              outcome   = normaldata$outcome, 
              complete  = normaldata$complete) 

out <- input %>%
  analysis(type = "normal")

str(out)
```

The trials is analyzed for the following hypothesis
$$H_0:\mu_{treatment} - \mu_{control} > 0 \quad H_A: \theta_{treatment} - \theta_{control} < 0$$
The delta and alternative used to analyze the trial is 0 and "less" respectively. 
The probability if accepting alternative is 0.95, the probability of stopping for futility is 0.05 and the probability of stopping for success is 0.90. 
The trial is reanalyzed with this input. 

```{r analysishypo}
input <- input %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") 

out <- input %>%
  analysis(type = "normal")

str(out)
  
```


The simulation is done for 40 times to impute the loss to follow up data and the posterior are sampled for 8000 (number_mcmc) to obtain an estimate for the posterior value.
The trial is reanalyzed again. 


```{r analysisimpute}
input <- input %>%
  impute(no_of_impute = 40, number_mcmc = 8000)

out <- input %>%
  analysis(type = "normal")

str(out)
```


Finally, there is no historical data incorporated for this trial. 
Instead of inputting parameters step by step, the trial parameters can be filled at once as illustrtaed below. 
The pipe function connects all input together and the analysis is done again. 

```{r analysisall}
out <- 
  data_normal(treatment = normaldata$treatment, 
              outcome   = normaldata$outcome, 
              complete  = normaldata$complete) %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  impute(no_of_impute = 40, number_mcmc = 8000) %>%
  analysis(type = "normal")

  
str(out)
```


