---
title: "Normal Outcome"
author: "Thevaa Chandereng, Donald Musgrove, Tarek Haddad, Graeme Hickey, Timothy Hanson, Theodore Lystig"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{bayesCT:normal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
set.seed(28999)
```

```{r, message = FALSE, echo = FALSE}
library(bayesCT)
```

# Introduction

The purpose of this vignette is to introduce how to simulate and analyze an adaptive bayesian clinical trial for normal outcomes. 
The simulation section compromise of the design of the trial itself which provides a type I error rate and power at each interim look. 
We use the normal-normal conjugate for the estimation of posterior probabilities. 
The historical data is used as an informative prior and we use `bayesDP` as an engine to incorporate the historical data. 


# Design


```{r}
value <-
  normal_outcome(mu_control = 16, mu_treatment = 13,
                 sd_treatment = 1.4, sd_control = 1.9) %>%
  historical_normal(mu0_treatment = 13, sd0_treatment = 2, N0_treatment = 100,
                    mu0_control = 12, sd0_control = 3, N0_control = 120, 
                    discount_function = "weibull", alpha_max = F, fix_alpha = 1,
                    weibull_scale = 0.135, weibull_shape = 3) %>%
  enrollment_rate(lambda = c(0.3, 1), time = 25) %>%
  study_details(total_sample_size = 300, study_period = 50,
                interim_look = c(210, 240, 270),
                prop_loss_to_followup = 0.10) %>%
  impute(no_of_impute = 25, number_mcmc = 1000) %>%
  randomize(block_size = 2, randomization_ratio = c(1, 1)) %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  simulate(no_of_sim = 5)


str(value)
```

# Analysis

```{r}
data(normaldata)

head(normaldata)
```



```{r}
value1 <-
  data_normal(data = normaldata) %>%
  analysis(type = "normal")

str(value1)
```


```{r}
value1 <-
  data_normal(data = normaldata) %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  analysis(type = "normal")

str(value1)
  
```


```{r}
value1 <-
  data_normal(data = normaldata) %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  impute(no_of_impute = 25, number_mcmc = 1000) %>%
  analysis(type = "normal")

str(value1)
```


```{r}

value1 <-
  data_normal(data = normaldata) %>%
  hypothesis(delta = 0, futility_prob = 0.05, prob_accept_ha = 0.95,
             expected_success_prob = 0.90, alternative = "less") %>%
  impute(no_of_impute = 25, number_mcmc = 1000) %>%
  historical_normal(mu0_treatment = 13, sd0_treatment = 2, N0_treatment = 100,
                    mu0_control = 12, sd0_control = 3, N0_control = 120, 
                    discount_function = "weibull", alpha_max = F, fix_alpha = 1,
                    weibull_scale = 0.135, weibull_shape = 3) %>%
  analysis(type = "normal")


str(value1)
```

